Testing the correlates of huge vertebrae number disparity in Australian blindsnakes

### 2024-10-09 manuscript.Rmd
addressed comments from Scott

---
title: "Dissociation between number of vertebrae and body size in Australian blindsnakes"
date: "Version `r Sys.Date()`"
author: ""
output: 
  # bookdown::html_document2:
  #   toc: true
  #   number_sections: FALSE
  bookdown::word_document2:
    reference_docx: [./custom-reference-doc.docx]
    number_sections: false
    fig_caption: yes
bibliography: [C:/Users/ST/Documents/repo/anu_thesis_texts/zotero_library.bib]
csl: evolution.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

@@

**Authors**: Sarin Tiatragul^1^; Andrew Amey^2^; Scott Hocknull^2^; Scott Keogh^1^; Emma Sherratt^3^

@@

<!-- **Author contributions**:  -->

<!-- ST and ES conceptualised the study; ST collected morphology data;  -->
<!-- ST obtained X-rays from SAM, WAM, and MAGNT. SH and AM provided X-rays Queensland Museum specimens. -->
<!-- ST and ES analysed the data; ST prepared the figures; ST, SK and ES wrote the manuscript. -->

<!-- **Acknowledgements (including funding information)**: -->

<!-- We to thank museum curators and collection managers from the  -->
<!-- South Australian Museum  (R Foster, S Donnellan), Western Australian Museum (P Doughty), Queensland Museum (A Amey and P Couper), Australian National Wildlife Collection (T Haff, C Wilson, and L Joseph), Museum and Art Gallery of the Northern Territory (G Dally), Museums Victoria (J Sumner, J Melville, and R Aguilar), Florida Museum of Natural History (J Gray, CM Sheehy, and ZS Randall), and the Natural History Museum Berlin (K Mahlow and F Tillack).  -->
<!-- SK thanks the Australian Research Council for ongoing support. ST PhD scholarship is supported by the Australian Government Research Training Program Scholarship at the Australian National University.  -->

@@

Keywords: Scolecophidia, vertebrae, pleomerism, snakes, axial

@@

Axial evolution is fundamental to vertebrate body plan evolution, facilitating body form changes and influencing modes of locomotion in mammals [@eg1500;@taewcharoenSmall2024;@galisFast2014]. reptiles [@eg1500;@vandammeSpeed2002;@molnarMorphological2023], amphibians [@eg1500;@bonettEvidence2017;@adlerEvolutionary2022] and fish [e.g., @yamadaAdaptive2009;@forkerBendy2021]. Variation of the axial skeleton is often studied by counting the number of vertebrae because the developmental mechanisms of segmentation are well characterised. Essentially vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryogenesis where the speed of this process relates to the size and number of somites, to produce what will become the axial skeleton [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009]. Elongate animals are expected to show pleomerism, where body size and number of vertebrae are positively correlated [@lindseyPleomerism1975]. Pleomerism has been observed in eels and fish [@wardEvolution2007;@wardDifferential2014;@reeceEvolutionary2013], in squamates [@pollyTesting2001;@headDissociation2007;@headEvolution2015;@bergmannVertebral2012;@lindellEvolution1994;@cernanskyVertebral2019], in salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015]. However, there are notable exceptions in some taxa [@headDissociation2007;@sherrattMacroevolution2022] indicating it is possible to disassociate segment production with somatic (body) growth, and revealing developmental processes underlying evolutionary patterns. Snakes are therefore a good model to study axial evolution because they exhibit great diversity in vertebrae number across species and have a wide range of body sizes and shapes without limb skeleton to confound signals of locomotor adaptation. Recent studies have indicated the number of vertebrae varies predictably with ecological diversification in elapid snakes [@sherrattHeterochronic2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024]. It also has important influence on snake locomotion, where the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997]. More research is required to understand the role that axial evolution plays in body plan and ecological diversification of snakes.

<!-- [Specific background]-->
Scolecophidians, commonly known in English as "blindsnakes" and "threadsnakes", are characterised by a reduced visual system and highly specialised phenotypes adapted for a strictly fossorial existence [@mirallesMolecular2018;@strongDeconstructing2021;@gowerEyetranscriptome2021]. Scolecophidians comprise five paraphyletic families, including the "core scolecophidians" — a monophyletic clade consisting of Typhlopidae (425 spp), Leptotyphlopidae (143 spp), Gerrhopilidae (29 spp), and Xenotyphlopidae (1 spp) — and the paraphyletic Anomalepididae (20 spp), which is sister to all other "typical snakes" (Alethinophidia) [@mirallesMolecular2018; @burbrinkInterrogating2020; @uetzQuarter2021]. 
Although all blindsnakes are fossorial, body size and shape vary significantly among species [@tiatragulMorphological2024; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993]. Leptotyphlopids are generally thought to be more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@herrelBurrowing2021]. These differences in body size and proportions determine how much push force the snakes can generate when burrowing. Their morphological differences led to the idea that leptyphlopids generally use existing burrows while typhlopids create their own burrow.

<!-- In another predominantly burrowing clade, caecilians, the number of vertebrae is not always correlated with body size [@renousBody1989], suggesting pleomerism is a mechanism for burrowing taxa to diversify in the underground environment.  -->

Pleomerism has been observed in typhlopidae, however, some notable exceptions indicate different mechanisms for gigantism [@headDissociation2007]. For instance, Head and Polly [-@headDissociation2007] found that some of the largest species of blindsnakes in the African continent (e.g. *Afrotyphlops schlegelii*) have fewer than expected precloacal vertebrae than expected for their size. This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae. Compared to other snakes, the vertebral column of fossorial snakes tends to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001]. This is reflected in the relatively short tails of blindsnakes [@tiatragulMorphological2024;@szyndlarIllustrated2023]. Limited studies on scolecophidian axial skeletons compared to alethinophidia means we have yet to appreciate the diversity that these groups have to offer. Given the recent shift in the perception of these earliest diverging clades of snakes (scolecophidians) away from the traditional view that they are "plesiomorphic and morphologically homogenous" [@strongConvergence2022], the scolecophidian axial skeleton also could offer a new perspective on the ecological origin of snakes.

The Australian blindsnake radiation comprise 47 described species [@tiatragulAustralia2023] distributed across all major biomes of Australia [@tiatragulPaleoenvironmental2023] with considerably diverse morphology from narrow-and-long (aspect ratio >= 100) to robust-and-wide (aspect ratio < 60). However, information about the relationship between body size and vertebrae number is unknown. In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses that deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes. The specific questions we addressed are: (1) How prevalent is pleomerism among Australian blindsnakes? (2) Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes? (3) Does vertebral ratio vary with soil compactness or mean annual temperature?

@@

We extracted 37 lineages of Australian blindsnakes from the time-calibrated phylogeny of Australian blindsnakes and outgroups presented in [@tiatragulAustralia2023]. The subset phylogeny represents 35/47 formally described species of _Anilios_ along with two additional genetically distinct lineages: one eastern population of "_A. grypus_" and one eastern population of "_A. ligatus_" [@tiatragulAustralia2023]. This phylogeny was inferred using the shortcut coalescent method in ASTRAL III [@zhangASTRALIIIPolynomialTime2018] based on 4930 maximum likelihood gene trees estimated using IQTREE [@minhIQTREE2020]. Divergence dating was estimated using MCMCTree from the PAML package [@yangPAML2007] based on 27 squamate reptile fossil calibrations [@tiatragulAustralia2023]. This phylogeny was used for all phylogenetic comparative analyses.

@@

We collected body size measurements and vertebral counts from museum specimens examined. To quantify body size and body width ratios, we measured snout-to-vent length (SVL), tail lengths (TL) and midbody diameter (MBD). We measured SVL and TL by stretching the specimen along a ruler to the nearest 0.5 mm. For MBD, we used a digital vernier calliper (ABS Coolant Proof Caliper, Mitutoyo, Japan) to the nearest 0.1 mm. Total body length (TBL) is the sum of SVL + TL. 
<!-- To obtain  mean body shape ratios [@mosimannSize1970] for each species, we divided each MBD by SVL and log-transformed for further analyses. -->

To obtain vertebral counts, we X-ray scanned the specimens at various institutions including: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n=87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n=129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (n=90), and Queensland X-Ray (n=35).

@@

We used the Fiji distribution of ImageJ2 v.1.54j [@schindelinFiji2012;@ruedenImageJ22017] to digitally visualise X-ray images and counted the number of vertebrae using the Cell Counter module. Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023]. We calculated vertebrae ratio by dividing the total number of vertebrae by total body length.

Additionally, we quantified intracolumnar length of each vertebra for three species representing (1) narrow-and-wide phenotypes (*A. grypus*), (2) wide-and-robust phenotypes (*A. pinguis*), and (3) intermediate phenotypes (*A. centralis*). Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 to place landmarks medially along the vertebral column at the anterior limit of the centrum of each vertebra. We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R Statistical Environment v.4.3.1 [@rcoreteamLanguage2023]. Measurements were repeated three times to account for measurement errors.

@@

From raster files that contained modelled approximations at the resolution of 3 arc-seconds, we extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html), mean annual temperature from WorldClim v.2 [@fickWorldClim2017] for each georeferenced record using the "extract" function in the R package *`raster`* v.3.6-3 [@hijmansRaster2022]. We then calculated the species maximum soil bulk density and species mean annual temperature.

@@

All statistical analyses were performed in R Statistical Environment. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in the R package *`geomorph`* v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum total body length from @tiatragulMorphological2024. Next, we test whether the vertebrae to body length ratio is related to the mean vertebrae number and mean aspect ratio, defined by total body length / body width, also from @tiatragulMorphological2024. Finally, we tested if variation in vertebrae ratio and vertebrae number can be explained by mean annual temperature and soil compactness. In all PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations. 
To visualise the relationship between vertebrae number and intracolumnar vertebral length, for each of the representative species, we plotted the mean vertebrae size (from three repeated measurements) against the vertebra number. Each vertebrae profile was overlaid with a fitted 4° polynomial curve.

@@

Total number of vertebrae varies among the 39 species of Australo-Papuan blindsnakes in this study (Fig. \@ref(fig:fig-tree-boxplots)), ranging from 146 in *A. tovelli* to 400 in *A. grypus* (**Table S1**). Aspect ratios ranged from 20.8 in the robust-and-wide *A. pinguis* to 169.8 in the narrow-and-long *A. longissimus* (**Table S2**).

```{r fig-tree-boxplots, fig.cap= 'Interspecific variation in aspect ratio and total number of vertebrae among Australo-Papuan blindsnake and close relatives. The boxplots represent the distribution of data, with vertical inside representing the median. The extent of the box represents the lower (left) and upper (right) quartile, with the whiskers showing 1.5 times the interquartile range. Outliers beyond this range are depicted as individual points. Some species are represented by only the mean.', out.width="165mm", out.height= "129mm"}

@@

For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, intercept=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig:fig-tbl-vert) and **Fig S1**). This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae. Many species have more vertebrae than expected for their size. For example, *A. grypus* "West" is an intermediate length (relative to other species) but has the greatest number of vertebrae, while *A. torresianus*, which is the longest, has relatively few vertebrae.

### Relationships between vertebrae ratio, number, and aspect ratio
We found that vertebrae ratio is positively correlated with aspect ratio (PGLS, slope=-0.0048, intercept=0.73, R^2^=0.17, F=7.09, Z=2.20, P=0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig. \@ref(fig:fig-vert-aspect-ratios)). We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope=-1.8, intercept=120.95, R^2^=0.69, F=79.39, Z=5.30, P<0.001).

<!-- Fig. Vertebral counts vs max body length  -->
```{r fig-tbl-vert, fig.cap= 'Maximum vertebral counts plotted against maximum total body length for Australian typhlopids. Trend line represent PGLS regression for Australian species. Phylogeny projected into space using maximum likelihood ancestral state estimation. Species not included in the phylogeny are shown in purple. ', out.width="165mm", out.height= "124mm"}

@@

<!-- Fig: Pleomerism and ratio correlations -->
```{r fig-vert-aspect-ratios, fig.cap= 'Species mean aspect ratios against (A) mean vertebrae ratio and (B) mean number of vertebrae of Australian blindsnakes. Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter. (C) Vertebral profile shape of three species representing different vertebrae ratios, and are depicted as a 4° polynomial curve with positive inflection. Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide. Phylogeny projected into space using maximum likelihood ancestral state estimation. Species not included in the phylogeny are shown in purple.', out.width="124mm", out.height= "165mm"}

@@

We found a positive relationship between vertebrae ratio and species mean annual temperature (PGLS, slope=0.038, y=0.19, R^2^=0.21, F=9.58, Z=2.32, P=0.009), indicating species that have greater number of vertebrae for body size are generally found in places that are warmer (Fig. \@ref(fig:fig-vert-ecology)A). We also found a significant negative relationship between vertebrae ratio and species maximum soil bulk density (PGLS, slope=-1.00, y=2.52, R^2^=0.12, F=4.8, Z=1.7, P=0.043), indicating species that have greater number of vertebrae for body size are generally found in places that have harder to penetrate soil (Fig. \@ref(fig:fig-vert-ecology)B). When considering raw mean vertebrae number, we found non-statistically significant positive correlations with both mean annual temperature (PGLS, slope=4.13, y=142.45, R^2^=0.76, F=2.87, Z=1.31, P=0.09) and maximum soil bulk density (PGLS, slope=111.69, y=, R^2^=0.04, F=1.61, Z=0.85, P=0.22; Fig. \@ref(fig:fig-vert-ecology)C–D).

<!-- Fig: Ratio and ecological variables -->
```{r fig-vert-ecology, fig.cap= 'Phylomorphospaces showing relationship between mean vertebrae ratios and mean vertebrae number against ecological variables. Mean vertebrae ratio against (A) species mean annual temperature and (B) species max soil bulk density. Mean vertebrae ratio is calculated by mean total vertebrae / mean total length for each species. Mean vertebrae number against (C) species mean annual temperature and (D) species max soil bulk density. Species mean annual temperature and max soil bulk density values are calculated based on occurrence data. Trend lines represent slopes from PGLS', out.width="165mm", out.height= "117mm"}

>>>>

---
# title: "Dissociation between number of vertebrae and body size in Australian blindsnakes"
title: "Testing the correlates of huge vertebrae number disparity in Australian blindsnakes"
date: "Version `r Sys.Date()`"
author: ""
output: 
  # bookdown::html_document2:
  #   toc: true
  #   number_sections: FALSE
  bookdown::word_document2:
    reference_docx: [./custom-reference-doc.docx]
    number_sections: false
    fig_caption: yes
bibliography: [C:/Users/ST/Documents/repo/anu_thesis_texts/zotero_library.bib]
csl: evolution.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

@@

**Authors**: Sarin Tiatragul^1^; Andrew Amey^2^; Scott Hocknull^2^; J. Scott Keogh^1^; Emma Sherratt^3^

@@

**CRediT author statement**:

ST: Conceptualisation, Methodology, Formal analysis, Writing - Original draft, reviewing and editing.
AM: Resources, Investigation
SH: Resources, Investigation
SK: Supervision, Funding acquisition, Writing - reviewing and editing.
ES: Conceptualisation, Methodology, Writing - reviewing and editing.

<!-- ST and ES conceptualised the study; ST collected morphology data -->
<!-- ST obtained X-rays from SAM, WAM, and MAGNT. SH and AM provided X-rays Queensland Museum specimens. -->
<!-- ST and ES analysed the data; ST prepared the figures; ST, SK and ES wrote the manuscript. -->

**Acknowledgements**:

We thank museum curators and collection managers from the 
South Australian Museum, Western Australian Museum, Queensland Museum, Australian National Wildlife Collection, Museum and Art Gallery of the Northern Territory, Museums Victoria, and the Florida Museum of Natural History.
SK thanks the Australian Research Council for ongoing support. ST PhD Scholarship was supported by the Australian Government Research Training Program Scholarship at the Australian National University.

@@

The evolution of the axial skeleton is important in understanding the evolution of body plans in vertebrates. Changes in the vertebral skeleton can influence body form and modes of locomotion. Elongate animals are commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism. However, some groups of vertebrates show a lack of pleomerism, suggesting it is possible to dissociate the production of segment and somatic growth, but what factors influenced this pattern is not well understood. Snakes offer a great model to study axial evolution because the number of vertebrae vary greatly between species and they exhibit a wide range of body size and shape. Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology can explain the variation in vertebrae number. We obtained X-ray scans and morphometric data for 42 described species of *Anilios* blindsnakes and evaluated their correlation with each other and ecological variables. We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebrae ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species. Finally, we found that species that have a greater number of vertebrae for their size are generally found in warmer environments and environments with harder-to-penetrate soil substrate. Our results extend our understanding of the axial evolution of basal snakes and the macroevolutionary dynamics of Australian blindsnakes.

Keywords: Scolecophidia, vertebrae, pleomerism, skeleton, axial

@@

Evolution of the axial skeleton is fundamental to vertebrate body plan evolution, facilitating body form changes and influencing modes of locomotion in mammals [e.g., @taewcharoenSmall2024;@galisFast2014], reptiles [e.g., @vandammeSpeed2002;@molnarMorphological2023], amphibians [e.g., @bonettEvidence2017;@adlerEvolutionary2022] and fish [e.g., @yamadaAdaptive2009;@forkerBendy2021]. Axial skeleton evolution is often studied by counting the number of vertebrae because the developmental mechanisms of segmentation are well characterised. Essentially vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryogenesis where the speed of this process relates to the size and number of somites, to produce what will become the axial skeleton [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009]. Elongate animals are expected to show a positive correlation between maximum adult body length and number of vertebrae, known as pleomerism [@lindseyPleomerism1975]. While pleomerism was considered the norm [@lindseyPleomerism1975;@wakeComparative1966;@pollyTesting2001], more recent studies have identified exceptions to pleomerism rule in moray eels [@reeceEvolutionary2013], large species of snakes [@headDissociation2007;@lindellEvolution1994], and salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].

The lack of pleomerism in some species suggests it is possible to disassociate segment production with body growth, and thus reveals the developmental processes underlying evolutionary patterns. Another clue to disassociation in reptiles can be found in the relationship between the number of ventral scales and precloacal vertebrae – a correlation that is 1:1 in many snakes [@alexanderPattern1966] and so ventral scales have often been used as a proxy for vertebrae number [@lindellEvolution1994;@shineVertebral2000]. However, this 1:1 correlation is absent in file snakes,*Achrochodus*, and some blindsnakes, Typhlopidae, [@alexanderPattern1966;@headDissociation2007], as well as some sea snakes of the genus *Hydrophis* [@vorisDermal1975], suggesting these taxa do not conform to pleomerism.
Snakes are therefore a good model to study axial evolution because they have great diversity in vertebrae number and a wide range of adult body sizes and shapes but without a limb skeleton to confound signals of locomotor adaptation. Recent studies have indicated the number of vertebrae varies predictably with ecological diversification in elapid snakes [@sherrattVertebral2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024]. 
The number of vertebrae has also been shown to influence snake locomotion, whereby the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997]. Despite these recent advances, there have been no study that have quantified the number of vertebrae across all the species in a diverse assemblage to test the prevalence of pleomerism in fossorial snakes.

<!-- [Specific background]-->
Scolecophidians, commonly known in English as "blindsnakes" and "threadsnakes", are characterised by a reduced visual system and highly specialised phenotypes adapted for a strictly fossorial existence [@mirallesMolecular2018;@strongDeconstructing2021;@gowerEyetranscriptome2021]. Scolecophidians comprise five paraphyletic families, including the "core scolecophidians" — a monophyletic clade consisting of Typhlopidae (425 spp), Leptotyphlopidae (143 spp), Gerrhopilidae (29 spp), and Xenotyphlopidae (1 spp) — and the paraphyletic Anomalepididae (20 spp), which is sister to all other "typical snakes" (Alethinophidia) [@mirallesMolecular2018; @burbrinkInterrogating2020; @uetzQuarter2021]. Although all blindsnakes are fossorial, body size and shape vary significantly among species [@tiatragulMorphological2024; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993]. Leptotyphlopids are generally considered more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@hedgesLower2008;@webbDietary2001;@webbPreysize1993;@iturriagaNew2021;@tiatragulMorphological2024]. These differences in body size and proportions determine how much push force the snakes can generate when burrowing, which led to the idea that leptyphlopids generally use existing burrows while typhlopids create their own burrow [@herrelBurrowing2021].

The relationship between segmentation and body elongation in scolecophidians is not well understood. However, a study by Head and Polly [-@headDissociation2007] that included 49 species across two genera typhlopids showed that vertebrae number and body size are positively correlated except when some of the largest species were included in the analysis. In particular, the largest species of blindsnakes in the African continent (*Afrotyphlops schlegelii*) have more precloacal vertebrae than expected for their size [@headDissociation2007]. This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae. Compared to other snakes, the vertebral column of fossorial snakes tends to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001], which is reflected in the relatively short tails of blindsnakes [@tiatragulMorphological2024;@szyndlarIllustrated2023]. Therefore, blindsnakes are a good system to test how changes in the number of pre-cloacal vertebrae affects maximum body length.

<!-- Limited studies on scolecophidian axial skeletons compared to alethinophidia means we have yet to appreciate the diversity that these groups have to offer. Given the recent shift in the perception of these earliest diverging clades of snakes (scolecophidians) away from the traditional view that they are "plesiomorphic and morphologically homogenous" [@strongConvergence2022], the scolecophidian axial skeleton also could offer a new perspective on the ecological origin of snakes.  -->
<!-- In another predominantly burrowing limbless clade, caecilian amphibians demonstrate a lack of pleomerism [@renousBody1989].  -->
<!-- Burrowing amphisbaenian lizards and some typhlopid blindsnakes have more ventral scales than vertebrae [@alexanderPattern1966] suggesting pleomerism. -->

The Australian blindsnake radiation, *Anilios*, comprise 47 described species distributed across all major biomes of Australia [@tiatragulAustralia2023;@tiatragulPaleoenvironmental2023]. Species in the genus show considerably diverse morphology from narrow-and-long to robust-and-wide [@tiatragulMorphological2024;@ehmannFamily1993]. However, information about the relationship between body size and vertebrae number is unknown. In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses about how deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes. The specific questions we addressed are: (1) How prevalent is pleomerism among Australian blindsnakes? (2) Can deviations from pleomerism be explained by body shape in Australian blindsnakes? (3) Does vertebral ratio vary with their ecology (e.g., soil compactness and mean annual temperature)?

@@

We extracted 37 lineages of Australian blindsnakes from the time-calibrated phylogeny of Australian blindsnakes and outgroups presented in @tiatragulAustralia2023. The subset phylogeny represents 35/47 formally described species of _Anilios_ along with two additional genetically distinct lineages: one eastern population of "_A. grypus_" and one eastern population of "_A. ligatus_" [@tiatragulAustralia2023]. This phylogeny was inferred using the shortcut coalescent method in ASTRAL III [@zhangASTRALIIIPolynomialTime2018] based on 4930 maximum likelihood gene trees estimated using IQTREE [@minhIQTREE2020]. Divergence dating was estimated using MCMCTree from the PAML package [@yangPAML2007] based on 27 squamate reptile fossil calibrations [@tiatragulAustralia2023]. This phylogeny was used for all phylogenetic comparative analyses.

@@

We collected body size measurements and vertebral counts from museum specimens examined. Body measurements including snout-to-vent length (SVL), tail lengths (TL) and midbody diameter (MBD) were obtained from @tiatragulMorphological2024. SVL and TL were measured to the nearest 1 mm on a ruler. MBD was measured using a digital vernier calliper (ABS Coolant Proof Caliper, Mitutoyo, Japan) to the nearest 0.1 mm. Total body length (TBL) is the sum of SVL and TL. To quantify body shape for each specimen, we calculated the aspect ratio [also referred to as elongation ratio; *sensu* @wardConvergent2004], which is the quotient of TBL and MBD.

To obtain vertebral counts, we X-ray scanned the specimens on the following machines and institutions: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n=87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n=129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (n=90), and Queensland X-Ray (n=35).

@@

We used the Fiji distribution of ImageJ2 v.1.54j [@schindelinFiji2012;@ruedenImageJ22017] to digitally visualise X-ray images and counted the number of vertebrae using the Cell Counter module. Pre- and post-cloacal vertebrae were distinguished based on vertebral morphology [@szyndlarIllustrated2023]. We calculated vertebrae ratio by dividing the total number of vertebrae by maximum TBL.

Additionally, we quantified intracolumnar length of each vertebra for three species representing (1) narrow-and-wide phenotypes (*A. grypus*), (2) wide-and-robust phenotypes (*A. pinguis*), and (3) intermediate phenotypes (*A. centralis*). Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 to place landmarks medially along the vertebral column at the anterior limit of the centrum of each vertebra. We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R Statistical Environment v.4.4.0 [@rcoreteamLanguage2024]. Measurements were repeated three times to account for measurement errors.

@@

We extracted soil bulk density values between 0 — 5cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html) and mean annual temperature from WorldClim v.2 [@fickWorldClim2017] from raster files that contained modelled approximations at the resolution of 3 arc-seconds. For each georeferenced record using the "extract" function in the R package *`raster`* v.3.6-3 [@hijmansRaster2022]. We then calculated the species maximum soil bulk density and species mean annual temperature.

@@

All statistical analyses were performed in R Statistical Environment v 4.4.0 [@rcoreteamLanguage2024]. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in the R package *`geomorph`* v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum TBL. Next, we tested whether vertebrae ratio is related to the mean vertebrae number and mean aspect ratio. Finally, we tested if variation in vertebrae ratio and vertebrae number is correlated with maximum soil bulk density and mean annual temperature. In all PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations. To visualise the relationship between vertebrae number and intracolumnar vertebral length, for each of the representative species, we plotted the mean vertebrae size (from three repeated measurements) against the vertebra number. Each vertebrae profile was overlaid with a fitted 4° polynomial curve [e.g., @sherrattMacroevolution2022].

@@

Total number of vertebrae varies greatly among the 39 species of Australo-Papuan blindsnakes in this study (Fig. \@ref(fig:fig-tree-boxplots)), ranging from 146 in *A. tovelli* to 400 in *A. grypus* (**Table S1**). Aspect ratios also vary greatly, ranging from 20.8 in the robust-and-wide *A. pinguis* to 169.8 in the narrow-and-long *A. longissimus* (**Table S2**). Intraspecific variation is also present in both aspect ratio and total number of vertebrae (Fig. \@ref(fig:fig-tree-boxplots)).

```{r fig-tree-boxplots, fig.cap= 'Interspecific variation in aspect ratio and total number of vertebrae among Australo-Papuan blindsnake and close relatives. The boxplots represent the distribution of data, with vertical thick line inside representing the median. The extent of the box represents the lower (left) and upper (right) quartile, with the whiskers showing 1.5 times the interquartile range. Outliers beyond this range are depicted as individual points. Some species are represented by only the mean.', out.width="165mm", out.height= "129mm"}

@@

For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum TBL (PGLS, slope=0.28, intercept=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig:fig-tbl-vert) and **Fig S1**). This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae. Many species have more vertebrae than expected for their size. For example, *A. grypus* "West" is an intermediate length (relative to other species) but has the greatest number of vertebrae, while *A. torresianus*, which has the greatest maximum TBL, has relatively few vertebrae.

### Relationships between vertebrae ratio, number, and aspect ratio
We found that vertebrae ratio is positively correlated with mean aspect ratio (PGLS, slope=-0.0048, intercept=0.73, R^2^=0.17, F=7.09, Z=2.20, P=0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig. \@ref(fig:fig-vert-aspect-ratios)). We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope=-1.8, intercept=120.95, R^2^=0.69, F=79.39, Z=5.30, P<0.001).

<!-- Fig. Vertebral counts vs max body length  -->
```{r fig-tbl-vert, fig.cap= 'Maximum vertebral counts plotted against maximum total body length (TBL) for Australian typhlopids. The trend line represents PGLS regression for Australian species. Phylogeny projected into space using maximum likelihood ancestral state estimation. Species not included in the phylogeny are shown in purple. ', out.width="165mm", out.height= "124mm"}

@@

<!-- Fig: Pleomerism and ratio correlations -->
```{r fig-vert-aspect-ratios, fig.cap= 'Species mean aspect ratios against (A) mean vertebrae ratio and (B) mean number of vertebrae of Australian blindsnakes. Vertebrae ratio is calculated by total vertebrae/TBL, and aspect ratio is TBL/MBD. (C) Vertebral profile shape of three species representing different vertebrae ratios, and are depicted as a 4° polynomial curve with positive inflection. Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide. Phylogeny projected into space using maximum likelihood ancestral state estimation. Species not included in the phylogeny are shown in purple.', out.width="124mm", out.height= "165mm"}

@@

We found a positive relationship between vertebrae ratio and species mean annual temperature (PGLS, slope=0.038, y=0.19, R^2^=0.21, F=9.58, Z=2.32, P=0.009), indicating species that have greater number of vertebrae for body size are generally found in places that are warmer (Fig. \@ref(fig:fig-vert-ecology)A). We also found a significant negative relationship between vertebrae ratio and species maximum soil bulk density (PGLS, slope=-1.00, y=2.52, R^2^=0.12, F=4.8, Z=1.7, P=0.043), indicating species that have greater number of vertebrae for body size are generally found in places that have  harder-to-penetrate soil (Fig. \@ref(fig:fig-vert-ecology)B). When considering raw mean vertebrae number, we found non-statistically significant positive correlations with both mean annual temperature (PGLS, slope=4.13, y=142.45, R^2^=0.76, F=2.87, Z=1.31, P=0.09) and maximum soil bulk density (PGLS, slope=111.69, y=, R^2^=0.04, F=1.61, Z=0.85, P=0.22; Fig. \@ref(fig:fig-vert-ecology)C–D).

<!-- Fig: Ratio and ecological variables -->
```{r fig-vert-ecology, fig.cap= 'Phylomorphospaces showing relationship between mean vertebrae ratios and mean vertebrae number against ecological variables. Mean vertebrae ratio against (A) species mean annual temperature and (B) species max soil bulk density. Mean vertebrae ratio is calculated by mean total vertebrae / mean TBL for each species. Mean vertebrae number against (C) species mean annual temperature and (D) species max soil bulk density. Species mean annual temperature and max soil bulk density values are calculated based on occurrence data. Trend lines represent slopes from PGLS', out.width="165mm", out.height= "117mm"}
